---
title: "EmbedSOM basic embedding"
author: "Mirek Kratochvil"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EmbedSOM basic embedding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
```

# Basic embedding with EmbedSOM

## Dataset

We will embed a small dataset created from gaussian clusters positioned in vertices of a 5-dimensional hypercube.

```{r}
#create the seed dataset
n <- 1024
data <- matrix(c(rep(0,n),rep(1,n)),ncol=1)

#add dimensions
for(i in 2:5) data <- cbind(c(rep(0,dim(data)[1]), rep(1, dim(data)[1])),rbind(data,data))

#scatter the points to clusters
set.seed(1)
data <- data + 0.2*rnorm(dim(data)[1]*dim(data)[2])
colnames(data) <- paste0('V',1:5)
```

This looks relatively nicely from the side (each corner in fact hides 8 separate clusters):
```{r, fig.show='hold'}
plot(data, pch=19, col=rgb(0,0,0,0.2))
```

Linear dimensionality reduction doesn't help much with seeing all 32 clusters:
```{r, fig.show='hold'}
plot(data.frame(prcomp(data)$x), pch='.', col=rgb(0,0,0,0.2))
```

Let's use the non-linear EmbedSOM instead.

## Getting the SOM ready

EmbedSOM works on a self-organizing map that you need to create first:

```{r}
set.seed(1)
map <- EmbedSOM::SOM(data, xdim=24, ydim=24)
```

EmbedSOM provides some level of compatibility with FlowSOM that can be used to simplify some commands. FlowSOM-originating maps and whole FlowSOM object may be used as well:

```{r eval=FALSE}
fs <- FlowSOM::ReadInput(as.matrix(data.frame(data)))
fs <- FlowSOM::BuildSOM(fsom=fs, xdim=24, ydim=24)
```

$24\times24$ is the recommended SOM size for getting something interesting from EmbedSOM -- it provides a good amount of detail, and still runs quite quickly.

## Embedding

When the SOM is ready, a matrix of 2-dimensional coordinates is obtained using the `EmbedSOM` function:

```{r}
e <- EmbedSOM::EmbedSOM(data=data, map=map)
```

Alternatively, FlowSOM objects are supported to be used instead of `data` and `map` parameters in most EmbedSOM commands:
```{r eval=FALSE}
e <- EmbedSOM::EmbedSOM(fsom=fs)
```

Several extra parameters may be specified; e.g. the following code makes the embedding a bit smoother and faster (but not necessarily better). See the EmbedSOM paper for details on parameters.

```{r}
e <- EmbedSOM::EmbedSOM(data=data, map=map, smooth=2, k=10)
```

Finally, `e` now contains the dimensionality-reduced 2D coordinates of the original data that can be used for plotting.

```{r}
head(e)
```

## Plotting the data

The embedding can be plotted using the standard graphics function, nicely showing all clusters next to each other.

```{r, fig.show='hold'}
plot(e, pch=19, cex=.5, col=rgb(0,0,0,0.2))
```

EmbedSOM provides specialized plotting function which is useful in many common use cases; for example for displaying density:

```{r, fig.show='hold'}
EmbedSOM::PlotEmbed(e, pch=19, cex=.5, nbin=100)
```

Or for seeing colored expression of a single marker (`value=1` specifies a column number; column names can be used as well):
```{r, fig.show='hold'}
EmbedSOM::PlotEmbed(e, data=data, pch=19, cex=.5, alpha=0.3, value=1)
```

(Notice that it is necessary to pass in the original data frame. When working with FlowSOM, the same can be done using `fsom=fs`.)

Or multiple markers:
```{r, fig.show='hold'}
EmbedSOM::PlotEmbed(e, data=data, pch=19, cex=.5, alpha=0.3, red=2, green=4)
```

Or perhaps for coloring the clusters. The following example uses the FlowSOM-style clustering to find the original 32 clusters in the scattered data. If that works right, each cluster should have its own color. (See FlowSOM documentation on how the meta-clustering works.)
```{r, fig.show='hold'}
n_clusters <- 32
hcl <- hclust(dist(map$codes))
metaclusters <- cutree(hcl,n_clusters)[map$mapping[,1]]

EmbedSOM::PlotEmbed(e, pch=19, cex=.5, clust=metaclusters, alpha=.3)
```

Custom colors are also supported (this is colored according to the dendrogram order):
```{r, fig.show='hold'}
colors <- topo.colors(24*24, alpha=.3)[Matrix::invPerm(hcl$order)[map$mapping[,1]]]

EmbedSOM::PlotEmbed(e, pch=19, cex=.5, col=colors)
```

`ggplot2` interoperability is provided using function `PlotGG`:
```{r, fig.show='hold'}
EmbedSOM::PlotGG(e, data=data) + ggplot2::geom_hex(bins=80)
```

(You may also get the ggplot-compatible data object using `PlotData` function.)

